\textbf{Machine learning classification by fitting amplicon sequences to existing OTUs}

### 2023-04-18 submission/manuscript.tex
second round revision with pat

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Machine learning classification using self-reference-based OTU clustering},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Machine learning classification using
self-reference-based OTU clustering}}
\author{}
\date{\vspace{-2.5em}}

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer. The signatures that associate with disease have
shown potential to aid in diagnosis of disease throught gut microbiome
profiling and machine learning. Taxonomic composition of microbial
communities can be assessed using amplicon sequencing of the 16S rRNA
gene, which is the input to classification models. Analysis of 16S rRNA
gene sequence data generally relies on assigning sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or \emph{de novo}. The quality of OTUs
generated with reference-based clustering is generally poor compared to
those generated with \emph{de novo} clustering
(\protect\hyperlink{ref-westcott2015}{1}). While \emph{de novo}
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{2}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene seequences to OTUs including, \emph{de novo},
database-reference-based, and self-reference-based with OptiFit. To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{3}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur 1 and 2) the VSEARCH algorithm used in
QIIME2 (\protect\hyperlink{ref-rognes2016}{4},
\protect\hyperlink{ref-bolyen2019}{5}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{1}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Previously, we observed that \emph{de novo}
clustering tends to produce higher MCC scores than reference-based
clustering (\protect\hyperlink{ref-sovacool2022}{2}). Consistent with
prior findings the reference-based methods produced lower MCC scores
(OptiFit Greengenes MCC = 0.786; VSEARCH Greengenes MCC = 0.531) than
the \emph{de novo} methods with the same algorithm (OptiClust \emph{de
novo} MCC = 0.884; VSEARCH \emph{de novo} MCC = 0.641). Another metric
we examined for the OptiFit workflow was the fraction of sequences from
the test set that mapped to the reference OTUs. Since sequences that did
not map to reference OTUs were eliminated, if a high percentage of reads
did not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the test set mapped to the reference OTUs. This number is higher than
the average fraction of reads mapped in the OptiFit Greengenes workflow
( 96.8\% +/- 3.5). These results indicate that the OptiFit
self-reference method performed as well as the OptiClust \emph{de novo}
method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the model performance was equivalent
between OptFit self-reference and OptiClust \emph{de novo} (p-value =
0.071; Figure 2A) while while performance for both VSEARCH methods was
lower than the OptiClust \emph{de novo}, OptiFit self, and OptiFit
Greengenes methods (p-values \textless{} 0.05). The trained model was
then applied to the test data classifying samples as either control or
SRN. Both VSEARCH methods perform slightly worse than the OptiClust *de
novo* method (both p-values \textless{} 0.05). However the performance
on the test data was equivalent between the OptiClust \emph{de novo},
OptiFit Greengenes, and OptiFit self-reference approaches (p-value
\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
data could be fit to existing OTU clusters using OptiFit without
impacting model performance.

@@

\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
were previously generated from human stool samples. Sequences were
downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
(\protect\hyperlink{ref-baxter2016}{3},
\protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
samples from 490 subjects. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'', while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
and 229 SRN samples.

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1`).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{2}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{9}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    The preprocessed data was fit to the clustered reference data using
    OptiFit with the method open to allow any sequences that did not map
    to the existing reference clusters would form new OTUs. The data was
    then subsampled to 10,000 reads and any novel OTUs from the test set
    were removed. The dataset was then split into two sets where 80\% of
    the samples were assigned to the training set and 20\% to the
    testing set. This process was repeated for each of the 100 splits
    resulting in 100 training and testing datasets.
  \end{enumerate}
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{5}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust 1 to generate OTUs. The data was subsampled to
  10,000 reads per sample and the resulting abundance tables were split
  into the training and testing sets. The process was repeated for each
  of the 100 splits resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{\emph{Machine Learning.}} A random forest model was trained with
the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{10}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{11}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

@@

The analysis workflow was implemented in Snakemake
(\protect\hyperlink{ref-koster2012}{12}). Scripts for analysis were
written in R (\protect\hyperlink{ref-R2020}{13}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{14}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{4}), RStudio
(\protect\hyperlink{ref-RStudio2019}{15}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{16}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{17}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{18}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{19}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
clustering methods outperform reference-based methods for assigning 16S
rRNA gene sequences to operational taxonomic units. PeerJ
\textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.}

\leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
\textbf{Mumphrey MB}, \textbf{Dotson GA}, \textbf{Schloss PD}. 2022.
OptiFit: An improved method for fitting amplicon sequences to existing
OTUs. mSphere \textbf{7}:e00916--21.
doi:\href{https://doi.org/10.1128/msphere.00916-21}{10.1128/msphere.00916-21}.}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Machine learning classification by fitting amplicon sequences to existing OTUs},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Machine learning classification by fitting amplicon
sequences to existing OTUs}}
\author{}
\date{\vspace{-2.5em}}

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential
diagnose of disease with gut microbiome sequence data and machine
learning. Taxonomic composition of microbial communities can be assessed
using amplicon sequencing of the 16S rRNA gene, which is the input to
classification models. Analysis of 16S rRNA gene sequence data generally
relies on assigning sequences into operational taxonomic units (OTUs).
The process of OTU clustering can either be reference-based or \emph{de
novo}. The quality of OTUs generated with reference-based clustering is
generally poor compared to those generated with \emph{de novo}
clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
novo} clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene sequences to OTUs including, \emph{de novo} and reference-based
clustering. For reference-based clustering, we used closed-reference
clustering to a public database (database-reference-based) and to OTUs
generated froma subset of the samples (self-reference-based). To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
and 2) the VSEARCH algorithm used in QIIME2
(\protect\hyperlink{ref-rognes2016}{7},
\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Consistent with prior findings, the reference-based
methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the test set mapped to the reference OTUs. This number is higher than
the average fraction of reads mapped in the OptiFit Greengenes workflow
( 96.8\% +/- 3.5). These results indicate that the OptiFit
self-reference method performed as well as the OptiClust \emph{de novo}
method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the performance of the OptFit
self-reference and OptiClust \emph{de novo} models were not
significantly different (p-value = 0.071; Figure 2A), while performance
for both VSEARCH methods was significantly lower than the OptiClust
\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
\textless{} 0.05). The trained model was then applied to the test data
classifying samples as either control or SRN. Both VSEARCH methods
perform slightly worse than the OptiClust \emph{de novo} method (both
p-values \textless{} 0.05). However the performance on the test data for
the OptiClust \emph{de novo}, OptiFit Greengenes, and OptiFit
self-reference approaches were not significantly different (p-value
\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
data could be fit to existing OTU clusters using OptiFit without
impacting model performance.

@@

\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
were previously generated from human stool samples. Sequences were
downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
(\protect\hyperlink{ref-baxter2016}{5},
\protect\hyperlink{ref-edgar2011}{9}). This dataset contains stool
samples from 490 subjects. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'', while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
and 229 SRN samples.

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{9}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust
  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
  fit to the clustered reference data using OptiFit with the method open
  to allow any sequences that did not map to the existing reference
  clusters would form new OTUs. The data was then subsampled to 10,000
  reads and any novel OTUs from the test set were removed. The dataset
  was then split into two sets where 80\% of the samples were assigned
  to the training set and 20\% to the testing set. This process was
  repeated for each of the 100 splits resulting in 100 training and
  testing datasets.
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
  generate OTUs. The data was subsampled to 10,000 reads per sample and
  the resulting abundance tables were split into the training and
  testing sets. The process was repeated for each of the 100 splits
  resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{\emph{Machine Learning.}} A random forest model was trained with
the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P-values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

@@

The analysis workflow was implemented in Snakemake
(\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis were
written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-Morgan2012}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Morgan XC}, \textbf{Tickle TL}, \textbf{Sokol
H}, \textbf{Gevers D}, \textbf{Devaney KL}, \textbf{Ward DV},
\textbf{Reyes JA}, \textbf{Shah SA}, \textbf{LeLeiko N}, \textbf{Snapper
SB}, \textbf{Bousvaros A}, \textbf{Korzenik J}, \textbf{Sands BE},
\textbf{Xavier RJ}, \textbf{Huttenhower C}. 2012. Dysfunction of the
intestinal microbiome in inflammatory bowel disease and treatment.
Genome Biol \textbf{13}:R79.}
